---
title: "Assessing the relationship between county-level stocks of community wealth and food system investment in the United States"
author: "Elliot Hohn"
format: html
editor: visual
---

## Introduction

The intersection of community wealth and United States Department of Agriculture's (USDA) support of local and regional food system in the form of grants represents a critical area of study for understanding the dynamics of local food systems. In recent years, the USDA has implemented various grant programs aimed at enhancing local and regional food markets, among which the Local Agriculture Market Program (LAMP) has been prominent. These grants are designed to support the development and expansion of local food enterprises, improve market opportunities for producers, and ultimately, enhance access to fresh, locally sourced food for consumers. The efficacy of such programs, however, may significantly depend on the socioeconomic fabric of the communities they aim to serve, particularly the wealth and resources available within these communities.

Community wealth, encompassing a range of economic indicators from median income and employment rates to more complex measures of economic diversity and resilience, provides a multifaceted view of the local economic environment. This wealth not only influences the immediate capacity of communities to engage with and benefit from USDA grants but may also affect the long-term sustainability and impact of the projects funded by such grants. Thus, exploring the relationship between community wealth metrics and USDA grant spending, especially under programs like LAMP, offers valuable insights into how federal funding can be optimized to support local food systems in diverse economic contexts.

This paper seeks to bridge the gap in the literature by conducting a comprehensive analysis of county-level community wealth metrics and their relationship with LAMP grant spending on local food systems. By employing a mix of descriptive statistics, spatial analysis, and cluster analysis, we aim to identify patterns and disparities in grant allocation and utilization across counties with varying levels of wealth. This analysis is particularly pertinent in the context of increasing recognition of the role local food systems play in promoting food security, economic development, and environmental sustainability.

Moreover, by focusing on LAMP grants, this study delves into a specific and impactful area of USDA's efforts to bolster local and regional food markets. These grants, which support a wide range of activities from infrastructure development to marketing and strategic planning for local food initiatives, serve as a critical tool for enhancing the agricultural economy at the local level. Understanding how these grants interact with the underlying community wealth will not only provide a clearer picture of the current landscape of federal support for local food systems but also offer insights into how such support can be tailored to meet the unique needs of different communities.

This paper contributes to the burgeoning field of agricultural economics and public policy by highlighting the nuanced ways in which federal grant programs intersect with local economic realities to shape the food landscape. Through a nuanced analysis of LAMP grant spending and community wealth metrics, we offer a comprehensive examination of the factors that contribute to the success and sustainability of local food systems initiatives, thereby informing future policy and program development in this vital sector.

## Methods

Plan right now is load county level community wealth data, and LAMP grant data, then perform cluster analysis on LAMP dataset

From [covid cluster analysis paper](https://www.biorxiv.org/content/biorxiv/early/2020/11/12/2020.11.12.379537.full.pdf): "This analysis was conducted in three stages. In Stage 0, county-level data were gathered from several sources, merged, and preprocessed for consistency. In Stage 1, time series clustering was performed on the number of newly reported confirmed COVID 19 cases per day (LAMP funding per year in our case?). Finally, in Stage 2, an explanatory model was fit to describe the relationship between cluster membership and several socioeconomic and political factors (community wealth factors?) describing the counties."

## Environment setup

```{r}
library(tidyverse)
library(sf)
library(openxlsx)
library(stringr)
library(corrplot)
library(stringr)
library(NbClust)
```

## Load community stocks data

Load data directly from CSU [github repo](https://github.com/CSU-Local-and-Regional-Food-Systems "CSU Local and Regional Food Systems repo").

```{r}
comm_stocks_url <- 'https://raw.githubusercontent.com/CSU-Local-and-Regional-Food-Systems/USDA-AMS-Data-and-Metrics/main/Indicators%20of%20Community%20Wealth/community_wealth.csv'
comm_stocks <- read.csv(comm_stocks_url)

# Function to add leading 0 to 4-digit fips values in comm_stocks
add_leading_zero <- function(x) {
  # Check if the number has 4 digits
  ifelse(nchar(x) == 4, paste0("0", x), x)
}

comm_stocks$fips <- sapply(comm_stocks$fips, add_leading_zero)
print(head(comm_stocks))
```

```{r}
# ggplot(comm_stocks, aes(x = value)) + 
#   geom_histogram(fill = 'red', colour = 'black') + 
#   facet_wrap(variable_name ~ .)

print(unique(comm_stocks$variable_name))
```

### Check correlations between variables

There are different years associated with the rows of data...how should I deal with this? Filter to only the newest data? Use the average? The grants data will also have a date associate with it, but I was thinking those grants would all be summed to get a "total amount invested" value.

I think for now I'm going to use the mean across all years for any given variable. A quick look at the data indicates that the values don't shift drastically from year to year, unsurprisingly.

```{r}
# reshape data from tall to wide to give each variable its own column
comm_stocks_wide <- comm_stocks %>%
  select(fips, year, variable_name, value) %>%
  group_by(fips, variable_name) %>%
  filter(fips != 0) %>%
  summarise(value = mean(value)) %>%
  ungroup() %>%
  pivot_wider(names_from = variable_name, values_from = value)
```

```{r}
# correlation matrix
correlation <- cor(comm_stocks_wide[, 2:ncol(comm_stocks_wide)], use = "complete.obs")
corrplot(correlation, method = 'square', tl.cex = 0.4, tl.col = 'black', tl.srt = 45,
         type = 'upper', order = "hclust")
```

### Ideas for additional variables

Rural/Underserved Counties: The Consumer Financial Bureau6 provides a list of rural or
underserved counties based on the Bureau’s interpretive rule (Regulation Z) using data from the
Home Mortgage Disclosure Act

### Check out some clustering approaches

First, trying out using the `NbClust` package to determine the optimal number of clusters.

```{r}
data <- na.omit(comm_stocks_wide[, 2:ncol(comm_stocks_wide)])
result <- NbClust(data, diss = NULL, distance = "euclidean", min.nc = 2, max.nc = 30, 
                  method = "ward.D2", index = "kl")

result$All.index
result$Best.nc
result$Best.partition

#kc <- kmeans(na.omit(comm_stocks_wide[, 2:ncol(comm_stocks_wide)]), 4)
```

Results are not great. It is currently not showing a lot of homogeneity across clusters, and not clear winner for optimal number of clusters.

### Load counties shapefile and join to community stocks data

```{r}
counties <- st_read('data/cb_2021_us_county_20m/cb_2021_us_county_20m.shp') %>%
  filter(!STATE_NAME %in% c('Alaska', 'Hawaii', 'Puerto Rico')) %>%
  select(GEOID, geometry) %>%
  st_transform(5070)

comm_stocks_geo <- merge(counties, comm_stocks_wide, by.x = 'GEOID', by.y = 'fips')

ggplot(comm_stocks_geo) +
  geom_sf(aes(fill = food_secure)) +
  scale_fill_distiller(
    name="Food Security", 
    palette = "BrBG",
    direction = 1
    ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2,"cm"),
        legend.key.height = unit(0.2,"cm")
  )
```

```{r}
summary_stats <- comm_stocks %>%
  group_by(variable_name) %>%
  summarize(
    mean = mean(value),
    median = median(value),
    sd = sd(value),
    max = max(value),
    min = min(value)
  )

print(summary_stats)
```

## Load LAMP data

Data downloaded directly from URL on the LAMP Navigator [webpage](https://www.ams.usda.gov/data/lamp-navigator "LAMP Navigator web pagef").

```{r}
lamp_url <- 'https://www.ams.usda.gov/sites/default/files/media/LAMPDatasetandDataDictionary.xlsx'

lamp <- read.xlsx(lamp_url, sheet = 'LAMP Dataset 2006 to 2023')
print(head(lamp))
```

### Load city lat/lon dataset

Loaded directly from [GitHub repo](https://github.com/kelvins/US-Cities-Database "city lat/lon data dump").

Going to try to join to cities from LAMP dataset.

```{r}
url <- 'https://raw.githubusercontent.com/kelvins/US-Cities-Database/main/csv/us_cities.csv'

city_loc <- read.csv(url)
```

Prep data for merge by renaming some columns, converting city names to all lower case in both data sets to ensure clean join. Note: there are some cities in the city_loc dataset that have more than one row, I think because they are bisected by county lines, so I dealt with that short list of issues by grouping them by city and state and selecting the first row in the group.

Also, there are three remaining rows in the LAMP data that don't have cities listed, so there are no coords for them, but I could track those down and fix them manually.

```{r}
lamp <- lamp %>%
  rename(
    state = State,
    city = City.2
    ) %>%
  mutate(city = str_squish(str_trim(tolower(city)))) %>%
  filter(!state %in% c('HI', 'AK', 'PR'))

city_loc <- city_loc %>%
  select(-c(ID, COUNTY)) %>%
  rename(
    city = CITY,
    state = STATE_CODE
  ) %>%
  mutate(city = str_squish(str_trim(tolower(city)))) %>%
  group_by(city, state) %>%
  filter(row_number()==1) %>% 
  ungroup()

# check how many cities in LAMP data are in city_loc data
print(length(unique(lamp$city)))
print(length(unique(lamp$city) %in% city_loc$city))

# do the join
lamp <- merge(lamp, city_loc, by = c('city', 'state'), all.x = TRUE)

# convert to sf object
lamp_sf <- lamp %>% 
  drop_na(LATITUDE, LONGITUDE) %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE")) %>%
  st_set_crs(4326) %>%
  st_transform(st_crs(comm_stocks_geo))
```

### Add LAMP grant points to the map

```{r}
ggplot(comm_stocks_geo) +
  geom_sf(aes(fill = food_secure)) +
  geom_sf(
    data = lamp_sf,
    size = 1.5,
    alpha = .5,
    color = 'purple',
    fill = 'purple'
    ) +
  scale_fill_distiller(
    name="Food Security", 
    palette = "BrBG",
    direction = 1
    ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2,"cm"),
        legend.key.height = unit(0.2,"cm")
  )
```

### Spatial join point data to county polygons

Need to assign lamp grant point data to county level

```{r}
# crs for each already aligns
lamp_by_county <- counties %>%
  st_join(lamp_sf, join = st_contains, left = TRUE) %>%
  group_by(GEOID) %>%
  mutate(total_grant = sum(Award.Amount))
```

### Look at distribution of funding

This should probably be per capita?

```{r}
pop_url <- 'https://www.ers.usda.gov/webdocs/DataFiles/48747/PopulationEstimates.csv?v=8269.9'

pop <- read.csv(pop_url) %>%
  filter(Attribute == 'CENSUS_2020_POP',
         FIPStxt != 0) %>%
  select(FIPStxt, Value) %>%
  rename(population_2020 = Value)

# re-use function from earlier to add leading zeros to FIPS
pop$FIPStxt <- sapply(pop$FIPStxt, add_leading_zero)

# clean up lamp data and add total county grant $ column before joining
lamp_county_summary <- lamp_by_county %>%
  select(-c("Organization", "Zip", "Project.Congressional.District", "Project.Title",
            "Project.Abstract/Goal", "Project.Start.Date", "Project.End.Date",
            "Not.Specified", "Project.Keywords", "StoryMap.Link", "Congressional.District",
            "Seed.Nof.Success.Highlights", "STATE_NAME", "Entity.Type", "Year", 
            "Unique.Project.Number", "Grant.Type", "Project.Type", "Project.Count",
            "Match.Amount.(Final)", "Total.Project.Cost", "total_grant"
            )) %>%
  drop_na() %>%
  summarise(across(Award.Amount:Value.Chain.Coordination, ~ sum(.x, na.rm = TRUE)))

lamp_county_summary <- lamp_county_summary %>%
  merge(pop, by.x = 'GEOID', by.y = 'FIPStxt') %>%
  mutate(award_per_capita = Award.Amount / population_2020) %>%
  mutate(centroid = st_centroid(geometry))

# change active geometry column to "centroid"
lamp_county_summary <- st_set_geometry(lamp_county_summary, "centroid")
```

Add points to map sized by per capita funding

```{r}
ggplot() +
  geom_sf(
    data = comm_stocks_geo,
    aes(fill = food_secure)) +
  geom_sf(
    data = lamp_county_summary,
    aes(size = award_per_capita),
    alpha = .5,
    color = 'purple',
    fill = 'purple'
    ) +
  scale_fill_distiller(
    name="Food Security", 
    palette = "BrBG",
    direction = 1
    ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2,"cm"),
        legend.key.height = unit(0.2,"cm")
  )
```

### Cluster LAMP grant county data

Need to figure out how to account for multiple years of grants, maybe? Or just sum it and use the total?

I'm going to start by summing all grants within a county and removing the time component...

```{r}
cluster_res <- lamp_county_summary %>% 
  select(-c(GEOID)) %>%
  st_drop_geometry() %>%
  NbClust(
    diss = NULL, distance = "euclidean", min.nc = 2, max.nc = 30, 
    method = "ward.D2", index = "kl")

cluster_res$All.index
cluster_res$Best.nc
membership <- cluster_res$Best.partition

# add cluster membership vector to 
lamp_county_summary$cluster_id <- membership
```

Add membership to map

```{r}
ggplot() +
  # geom_sf(
  #   data = comm_stocks_geo,
  #   aes(fill = food_secure)) +
  geom_sf(
    data = lamp_county_summary,
    aes(
      fill = cluster_id
    ),
    size = 1.5,
    alpha = .5#,
    # color = 'purple',
    # fill = 'purple'
    ) +
  # scale_fill_distiller(
  #   name="Food Security", 
  #   palette = "BrBG",
  #   direction = 1
  #   ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2,"cm"),
        legend.key.height = unit(0.2,"cm")
  )
```
