---
title: "Assessing the relationship between county-level indicators of community wealth and food system investment in the United States"
author: "Elliot Hohn"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
editor: visual
---

## Purpose

This markdown file is sort of a hybrid between an analytic script and a project report – it contains the narrative description of the project, methods, discussion, etc., but also all of the code for loading, cleaning, exploring, and analyzing the data, in addition to the code for creating plots, tables, etc. The reason for this is that I intend to solicit feedback from multiple people, some of whom will be interested in the code, and others who won't be interested in that at all. This format allows me to share the same output (pdf or html) with everyone to avoid separating the contextual information from the analysis itself.

In the end, this might get split into two components – one for the report and separate one for the data analysis. That's all TBD at this point.

## Introduction

The intersection of community wealth and support of local and regional food system in the form of federal grants represents a critical area of study for understanding the dynamics of local food systems. In recent years, the USDA has implemented various grant programs aimed at enhancing local and regional food markets, among which the Local Agriculture Market Program (LAMP) has been prominent. These grants are designed to support the development and expansion of local food enterprises, improve market opportunities for producers, and ultimately, enhance access to fresh, locally sourced food for consumers. The efficacy of such programs, however, may significantly depend on the socioeconomic fabric of the communities they aim to serve, particularly the wealth and resources available within these communities.

Community wealth, encompassing a range of economic indicators from median income and employment rates to more complex measures of economic diversity and resilience, provides a multifaceted view of the local economic environment. This wealth not only influences the immediate capacity of communities to engage with and benefit from USDA grants but may also affect the long-term sustainability and impact of the projects funded by such grants. Thus, exploring the relationship between community wealth metrics and USDA grant spending, especially under programs like LAMP, offers valuable insights into how federal funding can be optimized to support local food systems in diverse economic contexts.

This paper seeks to bridge the gap in the literature by conducting a comprehensive analysis of county-level community wealth metrics and their relationship with LAMP grant spending on local food systems. By employing a mix of descriptive statistics, spatial analysis, and cluster analysis, we aim to identify patterns and disparities in grant allocation and utilization across counties with varying levels of wealth. This analysis is particularly pertinent in the context of increasing recognition of the role local food systems play in promoting food security, economic development, and environmental sustainability.

Moreover, by focusing on LAMP grants, this study delves into a specific and impactful area of USDA's efforts to bolster local and regional food markets. These grants, which support a wide range of activities from infrastructure development to marketing and strategic planning for local food initiatives, serve as a critical tool for enhancing the agricultural economy at the local level. Understanding how these grants interact with the underlying community wealth will not only provide a clearer picture of the current landscape of federal support for local food systems but also offer insights into how such support can be tailored to meet the unique needs of different communities.

This paper contributes to the burgeoning field of agricultural economics and public policy by highlighting the nuanced ways in which federal grant programs intersect with local economic realities to shape the food landscape. Through a nuanced analysis of LAMP grant spending and community wealth metrics, we offer a comprehensive examination of the factors that contribute to the success and sustainability of local food systems initiatives, thereby informing future policy and program development in this vital sector.

## Methods

(TBD)

```{r environment setup}
#| include: false

library(tidyverse)
library(sf)
library(openxlsx)
library(stringr)
library(corrplot)
library(NbClust)

options(scipen = 999)
```

## Load and prepare "Indicators of Community Wealth" data

The original source of the community wealth data used in this analysis is from Schmitt et al. (2021). However, additional data processing has been done by researchers at Colorado State University (CSU) as part of the CSU Local and Regional Food Systems work, specifically the Indicators of Community Wealth data set included in the USDA-AMS-Data-and-Metrics [repository](https://github.com/CSU-Local-and-Regional-Food-Systems/USDA-AMS-Data-and-Metrics/tree/main) publicly available on GitHub (accessed March 20, 2024). Details on how the original [data](https://github.com/schmi-ny/County-Level-Community-Capital-Stocks/blob/main/CapitalPCs061521.xlsx) from Schmitt et al. (2021) were processed are described in an RMarkdown [file](https://github.com/CSU-Local-and-Regional-Food-Systems/USDA-AMS-Data-and-Metrics/blob/main/Indicators%20of%20Community%20Wealth/community_wealth.Rmd) within CSU's repository. The processed data is what is used in this analysis, loaded directly from the CSU GitHub repository.

After loading the data, a bit of exploratory data analysis is performed to better understand the data, then the values are normalized prior to performing a cluster analysis.

::: callout-note
The community wealth data has a year value associated with it, so there are often multiple values for a given metric for each county, each associated with a different year. I am not certain how to deal with this right now, so I'm calculating the mean of these values and removing the date.
:::

```{r community wealth data prep}
#| output: false

comm_wealth_url <- 'https://raw.githubusercontent.com/CSU-Local-and-Regional-Food-Systems/USDA-AMS-Data-and-Metrics/main/Indicators%20of%20Community%20Wealth/community_wealth.csv'
comm_wealth <- read.csv(comm_wealth_url) %>%
  filter(fips > 100) # remove national level rows

# Function to add leading 0 to 4-digit fips values in comm_wealth to facilitate later joins
add_leading_zero <- function(x) {
  # Check if the number has 4 digits
  ifelse(nchar(x) == 4, paste0("0", x), x)
}

# apply funcion to fips column
comm_wealth$fips <- sapply(comm_wealth$fips, add_leading_zero)

# make sure all fips codes are 5 digits
print(sum(nchar(comm_wealth$fips) != 5)) # should equal 0 if all fips codes are 5 digits
```

Show summary stats for all metrics in community wealth data.

```{r reshape community wealth and summarize}
#| warning: false
#| fig-height: 10

# reshape data from tall to wide to give each variable its own column
comm_wealth_wide <- comm_wealth %>%
  select(fips, year, variable_name, value) %>%
  group_by(fips, variable_name) %>%
  summarise(value = mean(value, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(names_from = variable_name, values_from = value)

print(summary(comm_wealth_wide))

comm_wealth_wide %>%
  pivot_longer(!fips, names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) + 
  geom_histogram(
    bins = 15
  ) + 
  facet_wrap(
    ~variable, 
    scales = 'free',
    ncol = 5,
    shrink = FALSE
    )
```

### Check correlations between variables

There are different years associated with the rows of data...how should I deal with this? Filter to only the newest data? Use the average? The grants data will also have a date associate with it, but I was thinking those grants would all be summed to get a "total amount invested" value.

I think for now I'm going to use the mean across all years for any given variable. A quick look at the data indicates that the values don't shift drastically from year to year, unsurprisingly.

```{r}
# correlation matrix
correlation <- cor(comm_wealth_wide[, 2:ncol(comm_wealth_wide)], use = "complete.obs")
corrplot(correlation, method = 'square', tl.cex = 0.4, tl.col = 'black', tl.srt = 45,
         type = 'upper', order = "hclust")
```

### Ideas for additional variables

Rural/Underserved Counties: The Consumer Financial Bureau6 provides a list of rural or underserved counties based on the Bureau’s interpretive rule (Regulation Z) using data from the Home Mortgage Disclosure Act

### Check out some clustering approaches

First, trying out using the `NbClust` package to determine the optimal number of clusters.

```{r}
data <- na.omit(comm_wealth_wide[, 2:ncol(comm_wealth_wide)])
result <- NbClust(data, diss = NULL, distance = "euclidean", min.nc = 2, max.nc = 30, 
                  method = "ward.D2", index = "kl")

result$All.index
result$Best.nc
result$Best.partition

#kc <- kmeans(na.omit(comm_wealth_wide[, 2:ncol(comm_wealth_wide)]), 4)
```

Results are not great. It is currently not showing a lot of homogeneity across clusters, and not clear winner for optimal number of clusters.

### Load counties shapefile and join to community stocks data

```{r}
counties <- st_read('data/cb_2021_us_county_20m/cb_2021_us_county_20m.shp') %>%
  filter(!STATE_NAME %in% c('Alaska', 'Hawaii', 'Puerto Rico')) %>%
  select(GEOID, geometry) %>%
  st_transform(5070)

comm_wealth_geo <- merge(counties, comm_wealth_wide, by.x = 'GEOID', by.y = 'fips')

ggplot(comm_wealth_geo) +
  geom_sf(aes(fill = food_secure)) +
  scale_fill_distiller(
    name="Food Security", 
    palette = "BrBG",
    direction = 1
    ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2,"cm"),
        legend.key.height = unit(0.2,"cm")
  )
```

```{r}
summary_stats <- comm_wealth %>%
  group_by(variable_name) %>%
  summarize(
    mean = mean(value),
    median = median(value),
    sd = sd(value),
    max = max(value),
    min = min(value)
  )

print(summary_stats)
```

## Load LAMP data

Data downloaded directly from URL on the LAMP Navigator [webpage](https://www.ams.usda.gov/data/lamp-navigator "LAMP Navigator web pagef").

```{r}
lamp_url <- 'https://www.ams.usda.gov/sites/default/files/media/LAMPDatasetandDataDictionary.xlsx'

lamp <- read.xlsx(lamp_url, sheet = 'LAMP Dataset 2006 to 2023')
print(head(lamp))
```

### Load city lat/lon dataset

Loaded directly from [GitHub repo](https://github.com/kelvins/US-Cities-Database "city lat/lon data dump").

Going to try to join to cities from LAMP dataset.

```{r}
url <- 'https://raw.githubusercontent.com/kelvins/US-Cities-Database/main/csv/us_cities.csv'

city_loc <- read.csv(url)
```

Prep data for merge by renaming some columns, converting city names to all lower case in both data sets to ensure clean join. Note: there are some cities in the city_loc dataset that have more than one row, I think because they are bisected by county lines, so I dealt with that short list of issues by grouping them by city and state and selecting the first row in the group.

Also, there are three remaining rows in the LAMP data that don't have cities listed, so there are no coords for them, but I could track those down and fix them manually.

```{r}
lamp <- lamp %>%
  rename(
    state = State,
    city = City.2
    ) %>%
  mutate(city = str_squish(str_trim(tolower(city)))) %>%
  filter(!state %in% c('HI', 'AK', 'PR'))

city_loc <- city_loc %>%
  select(-c(ID, COUNTY)) %>%
  rename(
    city = CITY,
    state = STATE_CODE
  ) %>%
  mutate(city = str_squish(str_trim(tolower(city)))) %>%
  group_by(city, state) %>%
  filter(row_number()==1) %>% 
  ungroup()

# check how many cities in LAMP data are in city_loc data
print(length(unique(lamp$city)))
print(length(unique(lamp$city) %in% city_loc$city))

# do the join
lamp <- merge(lamp, city_loc, by = c('city', 'state'), all.x = TRUE)

# convert to sf object
lamp_sf <- lamp %>% 
  drop_na(LATITUDE, LONGITUDE) %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE")) %>%
  st_set_crs(4326) %>%
  st_transform(st_crs(comm_wealth_geo))
```

### Add LAMP grant points to the map

```{r}
ggplot(comm_wealth_geo) +
  geom_sf(aes(fill = food_secure)) +
  geom_sf(
    data = lamp_sf,
    size = 1.5,
    alpha = .5,
    color = 'purple',
    fill = 'purple'
    ) +
  scale_fill_distiller(
    name="Food Security", 
    palette = "BrBG",
    direction = 1
    ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2,"cm"),
        legend.key.height = unit(0.2,"cm")
  )
```

### Spatial join point data to county polygons

Need to assign lamp grant point data to county level

```{r}
# crs for each already aligns
lamp_by_county <- counties %>%
  st_join(lamp_sf, join = st_contains, left = TRUE) %>%
  group_by(GEOID) %>%
  mutate(total_grant = sum(Award.Amount))
```

### Look at distribution of funding

This should probably be per capita?

```{r}
pop_url <- 'https://www.ers.usda.gov/webdocs/DataFiles/48747/PopulationEstimates.csv?v=8269.9'

pop <- read.csv(pop_url) %>%
  filter(Attribute == 'CENSUS_2020_POP',
         FIPStxt != 0) %>%
  select(FIPStxt, Value) %>%
  rename(population_2020 = Value)

# re-use function from earlier to add leading zeros to FIPS
pop$FIPStxt <- sapply(pop$FIPStxt, add_leading_zero)

# clean up lamp data and add total county grant $ column before joining
lamp_county_summary <- lamp_by_county %>%
  select(-c("Organization", "Zip", "Project.Congressional.District", "Project.Title",
            "Project.Abstract/Goal", "Project.Start.Date", "Project.End.Date",
            "Not.Specified", "Project.Keywords", "StoryMap.Link", "Congressional.District",
            "Seed.Nof.Success.Highlights", "STATE_NAME", "Entity.Type", "Year", 
            "Unique.Project.Number", "Grant.Type", "Project.Type", "Project.Count",
            "Match.Amount.(Final)", "Total.Project.Cost", "total_grant"
            )) %>%
  drop_na() %>%
  summarise(across(Award.Amount:Value.Chain.Coordination, ~ sum(.x, na.rm = TRUE)))

lamp_county_summary <- lamp_county_summary %>%
  merge(pop, by.x = 'GEOID', by.y = 'FIPStxt') %>%
  mutate(award_per_capita = Award.Amount / population_2020) %>%
  mutate(centroid = st_centroid(geometry))

# change active geometry column to "centroid"
lamp_county_summary <- st_set_geometry(lamp_county_summary, "centroid")
```

Add points to map sized by per capita funding

```{r}
ggplot() +
  geom_sf(
    data = comm_wealth_geo,
    aes(fill = food_secure)) +
  geom_sf(
    data = lamp_county_summary,
    aes(size = award_per_capita),
    alpha = .5,
    color = 'purple',
    fill = 'purple'
    ) +
  scale_fill_distiller(
    name="Food Security", 
    palette = "BrBG",
    direction = 1
    ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2,"cm"),
        legend.key.height = unit(0.2,"cm")
  )
```

### Cluster LAMP grant county data

Need to figure out how to account for multiple years of grants, maybe? Or just sum it and use the total?

I'm going to start by summing all grants within a county and removing the time component...

```{r}
cluster_res <- lamp_county_summary %>%
  select(-c(GEOID, geometry)) %>%
  st_drop_geometry() %>%
  NbClust(
    diss = NULL, distance = "euclidean", min.nc = 2, max.nc = 30,
    method = "ward.D2", index = "kl")

cluster_res$All.index
cluster_res$Best.nc
membership <- cluster_res$Best.partition

# add cluster membership vector to
lamp_county_summary$cluster_id <- membership
```

Add membership to map

```{r}
ggplot() +
  # geom_sf(
  #   data = comm_wealth_geo,
  #   aes(fill = food_secure)) +
  geom_sf(
    data = lamp_county_summary,
    aes(
      fill = cluster_id
    ),
    size = 1.5,
    alpha = .5#,
    # color = 'purple',
    # fill = 'purple'
    ) +
  # scale_fill_distiller(
  #   name="Food Security", 
  #   palette = "BrBG",
  #   direction = 1
  #   ) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2,"cm"),
        legend.key.height = unit(0.2,"cm")
  )
```
